Experience Report: Haskell in Computational Biology

### 2012-06-29 mrfy_experience_report.tex
lots of rewriting, much still in flight

@@

\newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{%
  \everypar=\expandafter{\the\everypar
       \verbatimsmallfalse
%       \belowdisplayskip=\abovedisplayskip
%       \belowdisplayshortskip=\abovedisplayshortskip
       \topsep=\standardvspace}%
  \topsep=0.78\topsep
%  \belowdisplayskip=0.78\belowdisplayskip
%  \belowdisplayshortskip=0.78\belowdisplayshortskip
  \verbatimsmalltrue\verbatiminput{#1}}
\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
\newcommand\smallfuzzverbatiminput[2]{{\par\unskip\small\hfuzz=#1 \verbatiminput{#2}}}

@@

\begin{document}
\parskip=0.9\parskip % cheating

@@

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallverbatiminput{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The entire right-hand side is wrapped in \texttt{extend~Mat}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
in~addition to computing the probability of the most likely path,
\mrfy\ also computes the most likely path (sequence of states).
Function \texttt{extend} adds~$M$ to the head of that path.

@@

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallverbatiminput{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with the search's age and the point's utility.
And the computation of \texttt{successors} should explain why we
separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
function that produces an infinite list:
We~don't \emph{want} the infinite list that would result from
applying \texttt{nextPt} many times in succession;
we~want the infinite lists that results from applying \texttt{nextPt}
to \texttt{startPt} many times in succession, each time with a
different source of randomness.

@@

\seclabel{perf}

At each point in its search,
MRFy calls \verb+vee'+ several
 times.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per point,
MRFy can keep six cores busy.
Given multiple placements per point,
it~can use forty-eight cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large\remark{QUANTIFY!!} protein structure, than when given 16 cores, which takes an average
of 37 seconds.

MRFy's predecessor, SMURF, does not perform stochastic search,
and thus its overall run-time performance cannot be compared directly to MRFy.
\remark{This \P\ is stultifying}
However, both programs implement the Viterbi algorithm.
On an exceedingly large, contrived example, comprising 56,000 amino acids
(larger than any known protein) and 343 nodes, and ignoring all beta-strand
information, SMURF's Viterbi implementation completed in 0.497 seconds on a
2.6GHz Core i7.
On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
seconds.
Despite the inferior runtime performance, MRFy still meets our performance
goals.
This is because the stochastic search algorithm relies on the Viterbi algorithm
applied to many, significantly smaller hidden Markov models than SMURF's
multidimensional programming must tackle.
MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
generally less than one minute, while SMURF consumes over 16GB of memory
and does not terminate even after eight hours.
Had we implemented the SMURF algorithm in Haskell, we believe that performance
would be unacceptable.
However, for the algorithm underlying MRFy, we are satisfied with the runtime
performance of Haskell and GHC.
Nonetheless, we expect that we can still improve MRFy's run-time performance
with further profiling and tuning.

In~\secref{viterbi} above, the example case for the \verb+vee'+
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{mrfy}, uses only the
score.
Even though Haskell evaluation is lazy, the \verb+vee'+ 
still allocates thunks that could compute paths.
We~cloned \verb+vee'+ and modified it to compute
only a score, with no path, and this change improved 
run-time performance by nearly~50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \verb+vee'+ that \emph{does} compute a path, so we can
extract the optimal parse for that placement.

How could we keep the performance improvement without having to
maintain two versions of \verb+vee'+?
In~Lisp or Ruby we would have used macros or metaprogramming,
but we were not confident of our ability to metaprogram a solution
using Template Haskell.
Instead, we replaced the primitive cons in \verb+vee'+ with a
call to a function of the same type, which we passed~in.
Passing the primitive~``\texttt{:}'' gives us a path,
and passing \verb+\_ _ -> []+ gives us the same
speedup as the cloned and modified~code.
But~although
this trick is simple and easy to implement, 
we~have an uneasy feeling that it may work only because of undocumented
properties of GHC's inliner, which may be subject to change without
notice.

\subsection{Awkward Haskell}

@@

We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
because we lack skills and knowledge in two areas.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and for most of MRFy's development, we used GHC 7.0, which
will happily add cost centers to top-level 
functions only, while much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose callgrind profiler gives useful information
without \emph{any} annotations in the source.
Late in MRFy's development, we began using GHC 7.4, which provides much
better profiling tools.
We are thrilled by these tool developments.

Our other ugly code came from  attempts to debug
run-time errors.
Our~group's legacy file
format is poorly documented: when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
These files are hard to deal with no matter what
programming language is used, and faults are to be expected.
But using Haskell, we had a hard time diagnosing run-time errors.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough:
the profiler can be used only on very small test cases that didn't always trigger
run-time errors.
This same limitation affected the GHCi debugger; our \texttt{vee'} function
shown in \secref{viterbi}, when it is not compiled with
optimizations, requires too much run time to be practical on non-trivial
inputs.
Moreover, the GHCi debugger can only set breakpoints at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

\remark{The following \P\ will need to be connected up to the last \S\
on ``our old approach to debugging.''}

Debugging led also not to awkward code, but to awkward interactions on
our team.
The two junior members of the team wanted to apply the debugging
skills they had honed through years of imperative programming.
The senior partner kept nattering on about QuickCheck properties.
No~progress was made until the first draft of this paper was prepared.
At~that point, we were quickly about to write QuickCheck
specifications for several properties of the Viterbi algorithm:
\begin{itemize}
\item
The total number of \verb+Mat+ and \verb+Ins+ states returned equals
the number of amino acids in the query sequence.
\item
The total number of matches, and deletions returned equals the
number of nodes in the hidden Markov model,
\emph{except} that an initial sequence of
of~\verb+Ins+ states counts as a single node,
due to an odd property of the HMMER file format (namely, the zeroth
node has no delete state and a non-emitting match state which
is always occupied, but a normal insert state).
\item
The sequence returned satisfies the Plan7 invariant, that is, states
\verb+Ins+ and \verb+Del+ are never adjacent.
\end{itemize}
\textbf{These properties need to be followed by more ambitious
properties},
such as proving that nearby sequences are less likely than the
sequence returned by \verb+vee'+.
One practice we recommend for those following in our footsteps is to
write instances of \texttt{Arbitrary} for data types as those data types
are written, when relevant invariants are fresh in memory.

This was a practice that the junior members of our team did not follow, and
they sorely regret it.
We speculate that one reason why we didn't follow this practice was a poor 
understanding of the \texttt{Arbitrary} type class.
Namely, a lot of the examples we had seen using QuickCheck focused 
predominantly on writing and testing properties with data types that already 
implemented \texttt{Arbitrary}.
We recall a smattering about \texttt{Arbitrary}, but we were still too baffled 
by the \texttt{Gen} monad to truly grasp its importance in using QuickCheck.
When it finally came time to use QuickCheck, we were overwhelmed by the amount 
of work that had to be done to write and test implementations of 
\texttt{Arbitrary} for each of our data types.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 1,200 lines of~Haskell.
It~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

In~\secref{awkward-profiling} above, we~mention the awkwardness of the
cost-center annotations required for profiling.
But~we also found it difficult just to get the profiler working.
In~particular, profiling requires that all installed libraries be recompiled
with profiling support.
We~were not able to achieve such recompilations consistently.

>>>>

\newcommand\primcons{\texttt{\small(:)}}

@@

\newenvironment{smallverbatim}{\par\small\verbatimsmalltrue\verbatim}{\endverbatim}
\newcommand\smallverbatiminput[1]{%
  \everypar=\expandafter{\the\everypar
       \verbatimsmallfalse
%       \belowdisplayskip=\abovedisplayskip
%       \belowdisplayshortskip=\abovedisplayshortskip
       \topsep=\standardvspace}%
  \topsep=0.78\topsep
%  \belowdisplayskip=0.78\belowdisplayskip
%  \belowdisplayshortskip=0.78\belowdisplayshortskip
  \verbatimsmalltrue\verbatiminput{#1}}
\newcommand\smallverbatiminputx[1]{\vspace*{-8pt}{\small\verbatiminput{#1}}} % WTF?? {\LaTeX} hell
\newcommand\smallfuzzverbatiminput[2]{%
  \everypar\expandafter{\the\everypar\hfuzz=0pt\relax}%
  \hfuzz=#1 \smallverbatiminput{#2}}

@@

\begin{document}
\parskip=0.9\parskip % cheating
\advance \parskip by 0pt plus 0.1pt

@@

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

@@

\seclabel{cons}
\seclabel{vee-prime}

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The right-hand side is wrapped in a call to
\texttt{\mbox{fmap (Mat `cons`)}}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ not only computes the probability of the most likely path but 
also computes the path itself.
Function \texttt{cons} adds~$M$ to a path;
we~avoid primitive~\primcons\ for reasons explained
in \secref{performance} below.

@@

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallfuzzverbatiminput{10.8pt}{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{costedUtility} to
tag each one with the search's age and the point's utility.
And the computation of \texttt{successors} should explain why we
separate \texttt{pt0}~from~\texttt{nextPt} instead of using a single
function that produces an infinite list:
We~don't \emph{want} the infinite list that would result from
applying \texttt{nextPt} many times in succession;
we~want the infinite lists that results from applying \texttt{nextPt}
to \texttt{startPt} many times in succession, each time with a
different source of randomness.

@@

\seclabel{perf}
\seclabel{performance}

At each point in its search,
MRFy calls \verb+vee'+ several
 times.
Our~\verb+vee'+
function computes a \mbox{\small \texttt{Scored [StateLabel]}}, that~is, 
an~optimal path and its score.
But at intermediate points in \mrfy's search,
\mrfy\ uses only the
score.
Even though Haskell evaluation is lazy, \verb+vee'+ 
still allocates thunks that could compute paths.
To~measure the relevant overhead,
we~cloned \verb+vee'+ and modified it to compute
only a score, with no path.
This change improved 
run time by nearly~50\%.
%%But when \texttt{search} finally decides on a placement, we need a
%%version of \verb+vee'+ that \emph{does} compute a path, so we can
%%extract the optimal parse for that placement.

How could we keep the improvement without 
maintaining two versions of \verb+vee'+?
In~Lisp or Ruby we would have used macros or metaprogramming,
but we were not confident of our ability to metaprogram a solution
using Template Haskell.
Instead, we use higher-order functions.
As~shown in \secref{vee-prime},
\verb+vee'+ does not use primitive~\primcons\ but instead
uses an unknown function \texttt{cons}, which is
passed~in.
To~get a path, we pass in primitive~\primcons;
to~get just a score, we pass in
\verb+\_ _ -> []+.
This trick is simple and easy to implement, and it provides the same
speedup as the cloned and modified~code.
But~we have an uneasy feeling that it may work only because of
undocumented properties of GHC's inliner, which may 
change without warning.

We~found another opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.\remark{Did NR break this?}
Even when there is just one placement per point,
MRFy can keep 6~cores busy.
\remark{This paragraph needs to talk about \textbf{speedups}.}
Given multiple placements per point,
it~can use 48~cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large\remark{QUANTIFY!!} protein structure, than when given 16
cores, which takes an average 
of 37 seconds.

MRFy's implementation of Viterbi's algorithm appears dramatically
slower than the C++ version implemented in \mrfy's predecessor,
SMURF.
On a 2.6GHz Intel Core i7,
given a contrived query sequence of 56,000 amino acids,
which is larger than any known protein,
with a hidden Markov model of 343 nodes and no beta strands,
\mrfy\ takes almost two minutes to compute an optimal parse.
SMURF parses the same sequence in half a second.
\remark{This result looks really bad. What is this contrived benchmark supposed to
mean, anyway??  Can we get a more realistic benchmark?
What fraction of time is spent in Viterbi?}
But unlike SMURF, \mrfy\ doesn't have to run on large models.
{\large{\textbf{QUANTIFY!}}}
\mrfy\ repeatedly runs Viterbi's algorithm on 
a~great many smaller models.
{\large{\textbf{QUANTIFY!}}}
Here, performance is good enough that \mrfy\ can do its job.
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;\remark{with or
without parallelism?}
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.
%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
%%% would be unacceptable.
%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
%%% performance of Haskell and GHC.
%
% Nonetheless, we expect that we can still improve MRFy's run-time performance
% with further profiling and tuning.  
% And we expect that if we cross the street, we will get to the other side.

\subsection{Awkward profiling and debugging}

@@

We've shown you some of the code we like, but we've also written 
code we dislike.
The~code we like least is what we've written to work around
obstacles we 
encountered in profiling and debugging.

GHC's profiler assigns costs not to functions but to ``cost
centers'' \cite{sansom-pj}.
In GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations can be added automatically only
to top-level 
functions.
But most of our code is in nested functions.
We~annotated such functions manually,
but the annotations
made our code so ugly that we felt compelled to
remove them as soon as possible.
Without the annotations, however, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose profiler gives useful information
without \emph{any} annotations.
(GHC~7.4 provides much
better profiling tools.
We~look forward to using them.)

We hate the code we've written to help debug
run-time errors.
Our~group's legacy file
format is poorly documented: when beta
strands overlap and are doubly paired, the invariants of the
format are not clear.
This format is hard to deal with no matter what
programming language is used, and faults are to be expected.
But using Haskell, we had a hard time diagnosing run-time errors.
Calls to
\texttt{trace} 
littered our code,
even when relegated to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, it didn't help:
the profiler can be used only on very small test cases that didn't always trigger
run-time errors.
This same limitation affected the GHCi debugger; 
in~GHCi, our \texttt{vee'} function
is too slow to be runnable on nontrivial
inputs.
Moreover, the GHCi debugger can only set breakpoints at top-level functions
or at specific column/line positions, which made debugging the memoized
\texttt{vee''} function impractical.
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

Our difficulties with debugging led to significant disagreements on
our team.
The two junior members wanted to apply the debugging
skills they had honed through years of imperative programming.
These skills did not transfer well to Haskell.
The senior member of the team kept saying that any proper approach to
debugging Haskell code should involve QuickCheck.
This advice was not helpful.

Our experience was that only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles to using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
For programmers who have seen lot of object-oriented dispatch
on \emph{arguments}, it is hard to grasp how the appropriate version
of \texttt{arbitrary} is chosen by Haskell's type-class system.
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
true importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did uncover a problem
using QuickCheck, we got witnesses that were much too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only because the whole team got involved in order to
meet the deadlines for this paper.

SUMMARY OF WHAT LITTLE WE HAVE LEARNED FROM QUICKCHECK
Have we tested that \verb+vee'+ returns a most likely sequence?
And didn't we find an inconsistency?

%%  At~that point, we were quickly about to write QuickCheck
%%  specifications for several properties of the Viterbi algorithm:
%%  \begin{itemize}
%%  \item
%%  The total number of \verb+Mat+ and \verb+Ins+ states returned equals
%%  the number of amino acids in the query sequence.
%%  \item
%%  The total number of matches, and deletions returned equals the
%%  number of nodes in the hidden Markov model,
%%  \emph{except} that an initial sequence of
%%  of~\verb+Ins+ states counts as a single node,
%%  due to an odd property of the HMMER file format (namely, the zeroth
%%  node has no delete state and a non-emitting match state which
%%  is always occupied, but a normal insert state).
%%  \item
%%  The sequence returned satisfies the Plan7 invariant, that is, states
%%  \verb+Ins+ and \verb+Del+ are never adjacent.
%%  \end{itemize}
%%  \textbf{These properties need to be followed by more ambitious
%%  properties},

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 1,200 lines of~Haskell.
It~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

In~\secref{awkward-profiling} above, we~mention the awkwardness of the
cost-center annotations required for profiling.
But~we also found it difficult just to get the profiler working.
In~particular, profiling requires that all installed libraries be recompiled
with profiling support.
Using Haskell's Cabal tool, which manages installed libraries,
we~were not able to achieve such recompilations consistently.

REWRITE THE ADVICE ON USING QUICKCHECK

One practice we recommend for those following in our footsteps is to
write instances of \texttt{Arbitrary} for data types as those data types
are written, when relevant invariants are fresh in memory.

This was a practice that the junior members of our team did not follow, and
they sorely regret it.
